---
title: "A short introduction to R"
author: "Boyko Amarov"
date: "29.10.2017"
output:
  html_document:
    self_contained: true
---

```{r, include=FALSE}
tutorial::go_interactive()
```

<div class="alert alert-danger">
This is a work in progress!
</div>

The purpose of this introduction is to give a short overview of R funcionality used in the exercise classes
accompanying the lecture on analysis and forecasting of economic time series. It is by no means intended to
serve as a reference or as an in-depth discussion of the lanuage itself.

# What is R and why use it?


R is an environment for data manipulation, analysis and data visualisation built around the R programming language.

At a first glance it may appear like a huge overkill to learn a programming language just
for the sake of doing some data analysis like fitting a regression model or estimating a mean. Other statistics
software like [Stata](https://www.stata.com/) and [SPSS](https://www.ibm.com/analytics/us/en/technology/spss/) offer some graphical user interfaces where a wide range of analysis can be performed simply by selecting options from menus. In my personal experience, however, doing any non-trivial analysis with those software packages eventually leads to writing some sort of script files (e.g. [Stata `.do` files](https://www.stata.com/manuals13/u16.pdf) or [SPSS syntax files](https://www.spss-tutorials.com/spss-syntax-editor-window/)) and for that you need to learn
those specific languages.


R is by no means the only environment that can be used for data analysis. Other languages with
support for data analysis are [Julia](https://julialang.org/) and [Python](https://www.python.org/).



# How to install R?


Windows users can download the latest installer from https://cran.r-project.org/bin/windows/base/.

GNU/Linux users can install it from the official repositories. For Ubuntu/Debian users:

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
sudo apt-get update && apt-get install r-base
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

MacOS users should follow the instructions on https://cran.r-project.org/.

## Packages ##

R consists of two parts: the base system and an ecosystem of packages contributed by the community. Apart from the core functionality available in base R, everything else is provided through community contributed packages packages.

As of 2017 there are more than 10,000 packages on CRAN (comprehensive R archive network).

A lot of packages in R are not self-contained but use functionality provided by other packages (dependencies). In order to install a package with dependencies, all the packages listed as dependencies must be installed beforehand.

To install a package (e.g. `ggplot2`) and all its dependencies, run:

~~~
install.packages('ggplot2')
~~~

To install a list of packages and all their dependencies (e.g. `ggplot2` and `zoo`), run:

~~~
install.packages(c('ggplot2', 'zoo'))
~~~

Most often we will use the functions and other objects provided in a package by loading the package into the environment. We will do it with the `library` function.

```{r}
# This will throw an error, (please look at the R console window to the right)
# Error: could not find function "qplot"
#because the functions in the
# ggplot2 package are not exposed in the workspace environment

qplot(rnom(50))

# We can call the qplot function directly from the workspace after
# loading the ggplot2 package that defines that function
library(ggplot2)

qplot(rnorm(50))

```

# How to start R?

The recommended way to interact with R for the purposes of this exercise classes is to use the [R-Studio](https://www.rstudio.com/products/rstudio/download/) desktop IDE. Free versions of R-Studio Desktop are
available for Windows, MacOS and for various flavours of Linux including Ubuntu/Debian, Fedora, RedHat and OpenSuse.

# R-Studio

[R-studio](https://www.rstudio.com/) is an integrated development environment (IDE) for the R language. While not strictly necessary for working with R files it makes makes working on R projects considerably less painful.

Please note that you must install R *before* launching R-Studio.

Furthermore, take a look at the R-Studio [cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf).



## Create a project with R-Studio

A R-Studio project is a convenient way to organise your work with R. In order to create a new project, start R-Studio and find the `File` menu in the upper left corner of the window. Select `New Project`, then choose one of
`New Directory` or `Existing Directory`. If you choose `New Directory`, click on `New Project` in the next step. Leave the options `create git repository` and `use packrat` blank and enter the name of the directory where you would like to create the project. Please use only ASCII symbols (latin letters) and `_` instead of blanks for the directory name.

## Create a script file

R script files are simple text files used to store a sequence of R commands. Per convention R script files use the `.R` extension. The `.R` file extension is important when working with R-Studio, because its R support features are enabled by default for `.R` files.

While you can use the R console directly to execute commands, script files are more convenient because you can reproduce the steps of the programme, track the files with a VCS (e.g. [git](https://git-scm.com/)) and share the files with other persons.


You can open and edit script files with any text editor but for non-trivial cases it is easier to use an editor with at least some support (syntax highlighting, linting, etc.) for the R language. Popular editors are [Emacs](http://www.gnu.org/software/emacs/) with the [ESS extension](http://ess.r-project.org/), [Vim](/http://www.vim.org/) with [Vim-R](http://www.vim.org/scripts/script.php?script_id=2628), [Eclipse](http://www.eclipse.org/eclipse/) with [StatET for R](http://www.walware.de/goto/statet), [Atom](/https://atom.io/) with [extensions](https://haroldpimentel.wordpress.com/2016/11/28/the-current-state-of-r-support-in-atom/) and [Tinn-R](https://sourceforge.net/projects/tinn-r/) among others.

In the exercise classes we will be using R-Studio.

When your project is successully created you can start using R by creating a script file. Open the `File` menu, click on `New File` and choose R-script. A new window named `Untitled1` should appear in the code pane. Click on the pane and save the file to disk by either clicking `File` -> `Save` or by pressing `Ctrl-S`. By default files are stored with `.R` extension.

When using a project, R-Studio configures R to use the project directory as a working directory. This is important because the working directory is used when resolving relative paths. If you use the R GUI directly, check the current working directory by executing:

~~~~~~~
getwd()
~~~~~~~

When writing R scripts or code in any language it is important to follow a consistent
style in order to produce *humanly* readable code. While there is no code standard for R,
you can take a look at some [guidelines](http://adv-r.had.co.nz/Style.html).

<!-- # Organising your work

## Git

While not strictly neccessary for working with R, a [version control system](https://en.wikipedia.org/wiki/Version_control) (VCS) can greatly ease the pain of managing collections of files.

There are may VCS solutions out there. Currently one of the most popular ones is [git](https://git-scm.com/).
Using R
------- -->

### A very simple case ###

In the simplest case we can use R as a sophisticated calculator. Note that lines starting
with `#` are comments. The R interpreter does not execute them. They are used to
document the programme and to give hints about the code in the file.
Open up a R console and type the following code:

```{r}
# Calculate the sum of 1 and 2
1 + 2

# Calculate the product of 4 and 7
4 * 7

#Divide 3 by 4
3 / 4

# Calculate 2 to the power of 8

3 ^ 8

# Calculate the (natural) logarithm of e
log(exp(1))

# Calculate the square root
sqrt(4)

# Take care with the precedence of operation
-2^2 == -4
(-2)^2 == 4

```

### Assigning objects ###

The previous example demonstrated some simple use of R for basic calculations.
We can take it one step further by assigning values to names. Use the `<-`
operator to assign a value to a name.

```{r}
# Calculate the sum of 1 and 2 to a variable called x
x <- 1 + 2

# Calculate the product of 4 and 7 to a variable called y
y <- 4 + 7

# Product of x * y = 3 * 11 = 33
x * y
```

Object names can contain letters, numbers and then chatacters `_` and '.' and must
start with a letter (names cannot contain blanks!). Choose the names for the objects you are creating carefully so that
they are both meaningful and short (not easy!). Ideally, a *person* reading your code should be able to understand it without running and debugging it. Separate words within a name with
'_' (`product_sales`) or use camel case style (`productSales`). Once you choose
a style, stick with it consistently throughout your code. Mixing styles generally
leads to confusing and hard to read code.

### Basic classes ###
<!-- TODO: finish! -->
Everything in `R` is an object. There are five basic (atomic) classes objects

* Character: for storing character strings like 'a', 'some string', etc.
* Numeric: for storing numbers
* Integer: for storing integer numbers
* Logical (TRUE/FALSE): for storing true/false values
* Complex

We can inspect the class of an object by calling the `class` function.
We can coerce objects to another basic class by using `as.character`, `as.numeric`, `as.logical` (if appropritate).

```{r}
x <- c(1.2, 3.2, 7.54)
class(x)

x_character <- as.charater(x)
x_character
class(charater)

y <- c(TRUE, FALSE, FALSE)
y
class(y)
as.numeric(y)
as.character(y)

# Beware that conversion from character to numeric
# is not always possible and will result in
# NA values (special values that represent missing data)

z <- c('1.3', '2', 'string')
as.numeric(z)
```


Logical operators
-----------------

Often we want to test if a value is less, greater of equal to another value.
This can be accomplished with the logical operators

* `<` less than
* `<=` less than or equal
* `==` exactly equal (*see the note below!*)
* `>` greater
* `>=` greater or equal
* `!=` not equal
* `!x` NOT x
* `x & y` *elementwise* x AND y: result has the length of the longer operand
* `x && y` logical AND: examines *only the first element of the operands*
* `x | y` *elementwise* x OR y: result has the length of the longer operand
* `x || y` logical OR: examines *only the first element of the operands*
* `is.na(x)` TRUE if x is `NA` (missing)
* `is.null(x)` TRUE if x is `NULL`

```{r}
1 < 2
2 <= 2
'some string' == 'some string'
'some string' == 'other string'
'some string' != 'other string'
!('some string' == 'other string')

TRUE && FALSE
TRUE && TRUE

FALSE || TRUE
FALSE || FALSE

x <- 1:5
x_less3 <- x < 3
x_less3
x_equal5 <- x == 5
x_equal5

x_less3 | x_equal5
x_less3 || x_equal5
x_less3 & x_equal5
x_less3 && x_equal5
```

**WARNING!** Take great care when comparing non-integer numbers. Read the [R FAQ](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f-1) and the references therein.

### Functions ###

A function is used to encapsulate pieces of code. A function is defined in the
following way.
```{r}
# Define a simple function and assign it to sum_xy
sum_xy <- function(x, y) {
  print(paste('X=', x, 'Y=', y))
  return(x + y)
}

# Call sum_xy with arguments x=1 and y=2. When calling the function with
# unnamed arguments their order MATTERS!
sum_xy(1, 2)
sum_xy(2, 1)

# The function can be called by explicitly passing the names of the arguments
sum_xy(x=1, y=2)
```


## Basic data structures ##

### Vectors ###

The basic data structure in R is called a vector. A vector is a (sequential) collection of
atomic data elements such as character strings, integers, numbers, etc. A vector is usually
created with the `c` (concatenate) function. A vector can
hold only elements of the same data type. If you try to create a vector with mixed
mixed character and numeric types R will try to coerce all elements to a common type.

```{r}
  # Create a vector x to store the numbers 1, 2, and 3
  x <- c(1, 2, 3)

  # Print the result
  x

  # You can concatenate vecotrs
  y <- c(x, 4, 5, 6);

  # WARNING: trying to create a vector with
  # mixed characters and number will coerce the numbers to characters
  mixed_vector <- c(1, 'some string', 2)
  mixed_vector
```

Arithmetic operations such as `+`, `-`, `*`, `^` are vectorised in the sense
that they are applied to each element of the vectors. The result of
using vectors of different lengths in an arithmetic operation has the
length of the longer vector. The shorter vector is recycled (see example).

```{r}
x <- c(1, 2, 3)
y <- x + 1
y
#This is the same as
z <- x + c(1, 1, 1)

# Take care when using vectors of different length
# The following creates a vector of integers from 1 to 4
v1 <- 1:4
v1
v2 <- 1:2
v2

# Adding vectors of different length
sum_v1v2 <- v1 + v2
sum_v1v2

# This is the same as explicitly repeating the shorter vector (v2)
# 2 times
v2_repeated <- rep(v1, 2)
v2_repeated
sum_v1v2_1 <- v1 + v2_repeated
sum_v1v2_1
```

## Factors ##
<!-- TODO: finish -->
Factors are a special class of vectors used for storing categorical variables.
Factors are useful for some regression models.

```{r}
x <- factor(c('male', 'female', 'male', 'female'))
x
levels(x)
```

## Inspecting an object ##

A couple of convinience functions are available to us for object inspection.
We have already seen the `class` funtion in the presentation of classes fo atomic data.

Another useful one is the `str` function

### Subsetting vectors ###
A very common operation with vectors is to choose a subset of a vector.

```{r}
x <- 1:4
x
#Select the first element of a vector
x[1]
#Select the last element of a vector:
x[length(x)]
#Select the first two elements of a vector
x[c(1, 2)]
#Select the first and third element of a vector
x[c(1, 3)]
# You can use a boolean vector as index
# Here we select the first and the last elements
x[c(TRUE, FALSE, FALSE, TRUE)]

#Negative index values are used to select every elemnet of the vector excluding
#those in the index

#Exclude the first element
x[-1]
#Exclude the last element
x[-length(x)]
```

## Data frames ##

Data frames are the basic structure that we will use to store data. You can view
a data frame as a table with columns usually corresponding to variables and
rows usually corresponding to observations. The columns in a data frame
(unlike in a matrix) need not be of the same type, i.e. we can have for example
numeric columns and character columns.

```{r}
df <- data.frame(x = 1:4, y = c('a', 'b', 'c', 'd'))
df
```

### Subsetting data frames ###

We can select subset of a data frame in a way very similar to the subsetting
of vectors by using the `[` function.

```{r}
df <- data.frame(x = 1:4, y = c('a', 'a', 'b', 'c'), z = rep(1:2, each = 2))

# Columns in df can be referred to with the $ function
# Select column x:

df$x

# Note: if the selected column does not exist, df$ will return NULL!

# Select the columns x and y, mind the comma in the bracket!
df[, c('x', 'y')]

#Select every column except x using a boolean vector to select columns
df[, !names(df) %in c('z')]

# Select the subset of df where column x is less than 3

df[df$x < 3, ]

# Select the subset of df where column y is equal to 'a'

df[df$y == 'a', ]

# Select the subset of df where column y == 'a' AND x > 1

df[df$x > 1 & df$y == 'a', ]
```

## Reading data ##

R supports reading data from a wide range of file formats. Data can also be accessed
directly from an URL. It also supports interfaces to a large number of database
management systems.

A very common format to store and exchange tabular data is CSV
(comma separated values). This is a plain text format where
each line in the file represents a record where fields are separated by some delimiter
character. Although the name CSV suggests that the separator should be a comma, CSV
is used to refer to a variety of formats and the delimiter can be almost any character,
including tabs, semicolons, colons, etc.

We will be using mainly two functions to read data from a plain text file: `read.csv`
and `read.table`. In the example below we read the data from a character string.
In most cases the data fill be stored in a file and `read.csv` would be called
with a file argument instead of a text argument. Both functions return a data frame.

```{r}
lines <- "
Year,Make,Model,Description,Price
1997,Ford,E350,some description,3000.00
1999,Chevy,Venture,another description,4900.00
"

cars <- read.csv(text=lines)
cars

# A useful function to inspect arbitrary objects
# Note that the column status is interpreted as a factor
str(cars)

# To change the default behaviour of read.csv and friends
# set the stringsAsFactor argument to FALSE

cars <- read.csv(text=lines, stringsAsFactor = FALSE)
str(cars)
```

## Getting help ##

The first source of help should be the R documentation. We can view the
built in documentation that comes with the R distribution by calling the
`help function`

~~~~
help(sum)
~~~~

or equivalently

~~~~
?sum
~~~~

Some packages include vignettes, a long form of documentation usually including example code with explanations. To list all vignettes included in a package (e.g. 'ggplot2'):

~~~
vignette(package = 'ggplot2')
~~~

The technical documentation is sometimes not that easy to understand or it can be hard to see how a particular problem can be solved only by reading the documentation.
A great place to start searching for answers is the R [mailing list archive](http://r.789695.n4.nabble.com/).

In those cases a google search can be the fastest way to find
a question that has already been answerd.

The [R FAQ list](https://cran.r-project.org/doc/FAQ/R-FAQ.html) list has collected a number of common problems and offers detailed explanations.

## Dates ##

Computer systems usually represent time as a number of ticks from
some origin. Internally, R stores time as the number of seconds from 01-01-1970
 and dates as the number of days from 01-01-1970. Use the `Date` class to store
 dates and the `POSIXct` or `POSIXlt` to store times. Use `as.Date` to convert
 character strings to `Date`. `as.Date` accepts a format argument where you need to declare the format of the string using a special syntax documented with the [`stptime`  function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/strptime.html). I can never remember all format specifiers and you should not try to learn those by heart either. Simply refer to the documentation:

~~~~~~~~
?stptime
~~~~~~~~


 For more information on how time values is handled in R, please refer to the [documentation](https://www.rdocumentation.org/packages/base/versions/3.4.1/topics/DateTimeClasses) and the references therein.

 We can create a character string from a date by using the `format` function and suppliyng it with the desired output format.


```{r}
# String in format monthe/day/year with century
date1 <- as.Date('01/03/2018', format = '%m/%d/%Y')
date1
format(date1, '%d %B %Y')

# String in format day-month-year with century
date2 <- as.Date('01-03-2018', format = '%d-%m-%Y')
date2
format(date2, '%d %B %Y')

# Don't rely on the defaults and always specify the format string
# unless you absolutely know what you are doing
date3 <- as.Date('01/03/2018')
format(date3, '%d %B %Y')


# as.Date will throw an error if it encounters invalid date strings
as.Date('01/invalid month/2018')
 ```

Take care when formatting dates and remember that some format specifications (e.g. '%B') are locale specific! Locale specific means that if R is using the 'en_US' locale, the result of `%B` for January will be `January` but for a user with a R using the 'bg_BG' locale the result of the same call will be `януари`.

## Basic summaries ##


# Graphics #

We will be using the `ggplot2` library for data visualisations.
